Biden Predicted to Secure a Minimum 10% Lead in Popular Vote Over Trump in the 2024 Election

### 2024-03-16 outputs/paper.qmd
final :D

I will use multilevel regression with post-stratification (MRP) to model and then predict whether Biden or Trump will win the popular vote, which is the total number of votes cast by individual citizens. MRP in this paper involves collecting individual-level survey data from the Polarization Research Lab [@polarization], analyzing the proportion of support among different demographical groups, and applying the proportion to a larger census dataset from the IPUMS USA, 2022 American Community Survey [@ipumsIPUMS]. One thing needs to be pointed out is that rather than only depending on the popular vote, the United States elects its president though electoral college. The 51 states are allocated 538 electoral-college votes. The numbers of votes granted to each state, which are different, depend on its representation in the Senate and House of Representatives, and usually every state gives all of its electoral-college votes to whomever wins the popular vote in that state [@economistDoesAmericas]. In other words, it is possible that one candidate wins the popular vote, representing the collective choice of voters, but loses the presidential election because of the distribution of electoral college votes, like the election in 2016 where Hillary Clinton won the popular vote (48%) while losing the election because the states that supported her had fewer electoral college votes than those supporting Trump [@nytimes2016Presidential].

@@

The data utilized for analysis, including both survey and census data, is shown in @sec-data. Following that, @sec-model introduces the model designed to generalize patterns and forecast election outcomes. I will also talk about the predicted results derived from the model in @sec-results. Lastly, @sec-discussions contains the results in a broader context and address weaknesses of this paper as well as the future explorations. This paper uses the programming language R [@r]. The analysis, the model and the visualizations use the following packages: ipumsr [@ipumsr], dplyr [@rdplyr], readr [@rreadr], tidyr [@rtidyr], arrow [@rarrow], ggplot2 [@rggplot2], knitr [@rknitr], statebins [@rstatebins], modelsummary [@rmodelsummary], broom [@rbroom], here [@rhere], and kableExtra [@rkableExtra].

@@

\newpage
# Data {#sec-data}
## Survey Data and Variables Selection
The survey I use is from the America’s Political Pulse Survey by the Polarization Research Lab [@polarization]. This open-access online survey contains a series of core questions on politics and special questions designed by political science researchers, posing inquiries to one thousand Americans every week. All the respondents, the population, are paid survey takers from the YouGov survey platform. It collects demographic information, party each respondent support, and core questions like Democrat/Republican thermometer or pride in being American.

The datasets are available weekly. For this paper, I choose 20 weeks of data, spanning from October 6th, 2023 to February 23rd, 2024, forming a dataset with 20,000 observations and 69 variables. The variable ‘pid3’ shows the answers to ‘Generally speaking, do you think of yourself as a…?’, with responses being Democratic, Republican, or Independent. For this paper, I will use this variable as supporting the Democratic Party/Biden or Republican Party/Trump.

In order to predict the 2024 presidential election result using several predictors, I first need to choose variables that help effectively differentiate between Democrats and Republicans. The survey dataset results show that more females support the Democratic Party than males, with 63.9% female and 58.2% male respondents supporting Democracy. People with higher levels of education tend to support the Democratic Party, especially those who completed post-graduate studies. 55.1% of high school graduates and 60.1% of people with no high school support the Democratic Party, while 72.8% of post-grads support it. Widowed and married individuals lean Republican, while divorced, never married, and separated people lean Democratic. Ethnicities exhibit varied preferences; around 74.8% of Asians and 79.7% of African Americans support Biden, compared to 55.7% of white individuals. Additionally, since the presidential election is voted by state, I would select race, gender, marital status, highest level of education and state of residence as variables to explain the binary support for Biden. Details about data cleaning can be found in @sec-appendix-cleaning.

@@

## Post-Stratification Data
IPUMS USA is a website that contains decennial census from 1790 to 2010, as well as survey data from the American Community Surveys (ACS) and the Puerto Rican Community Surveys from 2000 to the present [@ipumsIPUMS]. The annual ACS public-use samples typically include housing data, demographic data, economic data, and some other individual characteristics. For this analysis, I will use the most recent American Community Survey, the 2022 Sample, as the post-stratification dataset to represent the voter population. This survey used cluster sampling to obtain 1-in-100 national random sample of the population. The population for this sample was the entire US population, categorized into households as subgroups, and the responses of the sampled subgroups were collected.

@@

[@fig-population_map] is a state bin plot of the predicted support for Joe Biden and Donald Trump by states in a gradient colour scale. The colour intensity indicates the level of support. The bluer a state is, the higher the predicted proportion of support for Biden. The redder a state appears, the higher the predicted proportion of support for Trump. Once again, it appears that a greater number of states prefer Biden over Trump, with an overall 62.96% supporting Biden.

@@

\newpage
# Discussions {#sec-discussions}
As noted in the introduction, the outcome of the presidential election depends on electoral college votes, and unexpected outcome, like the one in 2016, may occur. Whether using the balanced or unbalanced datasets, both suggest a substantial lead for Biden in the popular vote, surpassing 10%. This significant advantage makes it highly likely for him to secure more electoral college votes and thus, ensure his re-election as the President of the United States.

>>>>

I will use multilevel regression with post-stratification (MRP) to model and then predict whether Biden or Trump will win the popular vote, which is the total number of votes cast by individual citizens, with the estimand being the proportion of voters supporting each candidate. MRP in this paper involves collecting individual-level survey data from the Polarization Research Lab [@polarization], analyzing the proportion of support among different demographical groups, and applying the proportion to a larger census dataset from the IPUMS USA, 2022 American Community Survey [@ipumsIPUMS]. One thing needs to be pointed out is that rather than only depending on the popular vote, the United States elects its president though electoral college. The 51 states are allocated 538 electoral-college votes. The numbers of votes granted to each state, which are different, depend on its representation in the Senate and House of Representatives, and usually every state gives all of its electoral-college votes to whomever wins the popular vote in that state [@economistDoesAmericas]. In other words, it is possible that one candidate wins the popular vote, representing the collective choice of voters, but loses the presidential election because of the distribution of electoral college votes, like the election in 2016 where Hillary Clinton won the popular vote (48%) while losing the election because the states that supported her had fewer electoral college votes than those supporting Trump [@nytimes2016Presidential].

@@

The data utilized for analysis, including both survey and census data, is in @sec-data. Following that, @sec-model introduces the model designed to generalize patterns and forecast election outcomes. I will also talk about the predicted results derived from the model in @sec-results. Lastly, @sec-discussions contains the results in a broader context and address weaknesses of this paper as well as the future explorations. This paper uses the programming language R [@r]. The analysis, the model and the visualizations use the following packages: ipumsr [@ipumsr], dplyr [@rdplyr], readr [@rreadr], tidyr [@rtidyr], arrow [@rarrow], ggplot2 [@rggplot2], knitr [@rknitr], statebins [@rstatebins], modelsummary [@rmodelsummary], broom [@rbroom], here [@rhere], and kableExtra [@rkableExtra].

@@

\newpage
# Data {#sec-data}
## Survey Data and Variables Selection
The survey I use is from the America’s Political Pulse Survey by the Polarization Research Lab [@polarization]. This open-access online survey contains a series of core questions on politics and special questions designed by political science researchers, posing inquiries to one thousand Americans every week. All the respondents, the population, are paid survey takers from the YouGov survey platform. The survey collects demographic information, party affiliation of each respondent, and core questions such as Democrat/Republican thermometer or pride in being American. It aims to illuminate public sentiment towards political animosity, respect for democracy, and support for political violence [@polarization].

The datasets are available weekly. For this paper, I choose 20 weeks of data, spanning from October 6th, 2023 to February 23rd, 2024, forming a dataset with 20,000 observations and 69 variables. The variable ‘pid3’ shows the answers to ‘Generally speaking, do you think of yourself as a…?’, with responses being Democratic, Republican, or Independent. Since there is no direct question asking respondents whom they are planning to vote for in the upcoming election, I will use this variable 'pid3' to indicate support for the Democratic Party/Biden or the Republican Party/Trump in this paper. 
In order to predict the 2024 presidential election result using several predictors, I first need to choose variables that help effectively differentiate between Democrats and Republicans. The survey dataset results show that more females support the Democratic Party than males, with 63.9% female and 58.2% male respondents supporting Democracy. People with higher levels of education tend to support the Democratic Party, especially those who completed post-graduate studies. 55.1% of high school graduates and 60.1% of people with no high school support the Democratic Party, while 72.8% of post-grads support it. Widowed and married individuals lean Republican, while divorced, never married, and separated people lean Democratic. Ethnicities exhibit varied preferences; around 74.8% of Asians and 79.7% of African Americans support Biden, compared to 55.7% of white individuals. Additionally, since the presidential election is voted by state, I would select race, gender, marital status, highest level of education and state of residence as variables to explain the binary support for Biden. Details about data cleaning can be found in @sec-appendix-cleaning.

@@

## Post-Stratification Data
IPUMS USA is a website that contains decennial census from 1790 to 2010, as well as survey data from the American Community Surveys (ACS) and the Puerto Rican Community Surveys from 2000 to the present [@ipumsIPUMS]. The ACS employed a monthly rolling sample of households, designed to replace the census long forms. The sampling unit was the household and encompassed all persons residing within.The annual ACS public-use samples typically include housing, demographic and economic data, as well as some other individual characteristics. For this analysis, I will use the most recent American Community Survey, the 2022 Sample, as the post-stratification data to represent the voter population. This is a weighted sample which includes persons in group quarters. This survey used cluster sampling to obtain 1-in-100 national random sample of the population. The population for this sample was the entire US population, categorized into households as subgroups, and the responses of the sampled subgroups were collected [@ipumsIPUMS].

@@

[@fig-population_map] is a state bin plot of the predicted support for Joe Biden and Donald Trump by states in a gradient colour scale. The colour intensity indicates the level of support. The bluer a state is, the higher the predicted proportion of support for Biden. The redder a state appears, the higher the predicted proportion of support for Trump. Once again, it appears that a greater number of states prefer Biden over Trump, with an overall 62.96% supporting Biden and 37.04% supporting Trump.

@@

\newpage
# Discussions {#sec-discussions}
Using MRP and post-stratification, I predict that Biden will have a minimum 10% lead in popular vote during the upcoming 2024 Presidential Election. As noted in the introduction, the outcome of the presidential election depends on electoral college votes, and unexpected outcome, like the one in 2016, may occur. Whether using the balanced or unbalanced datasets, both suggest a substantial lead for Biden in the popular vote, surpassing 10%. This significant advantage makes it highly likely for him to secure more electoral college votes and thus, ensure his re-election as the President of the United States.
